---
title: "Getting Started with ClinicoPathDescriptives"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Getting Started with ClinicoPathDescriptives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

ClinicoPathDescriptives is a comprehensive R package designed for descriptive analysis in clinicopathological research. It provides both programmatic R functions and a graphical user interface through jamovi, making advanced statistical analysis accessible to researchers with varying levels of programming experience.

### Key Features

- **Descriptive Statistics**: Generate comprehensive "Table One" summaries
- **Cross-tabulations**: Create contingency tables with appropriate statistical tests  
- **Visualization Tools**: Age pyramids, alluvial diagrams, Venn diagrams, variable trees, and more
- **Data Quality Assessment**: Benford's law analysis, outlier detection, single- and multi-variable quality checks
- **Dual Interface**: Use in R or through jamovi's point-and-click interface

> **Note on treatment-response plots.** Waterfall and swimmer plots are *not* part of
> ClinicoPathDescriptives. They are provided by the **OncoPath** module
> (`Analyses > OncoPath > Patient Follow-Up Plots`).

## Installation

### Install from GitHub

```{r eval=FALSE}
# Install development version
if (!require("remotes")) install.packages("remotes")
remotes::install_github("sbalci/ClinicoPathDescriptives")
```

### Install as jamovi Module

Download the `.jmo` file from [Releases](https://github.com/sbalci/ClinicoPathDescriptives/releases) and install in jamovi via: `jamovi menu > Modules > Install from file`

## Loading the Package

```{r setup}
library(ClinicoPathDescriptives)
library(dplyr)    # For data manipulation
library(ggplot2)  # For additional plotting
```

## Example Dataset

The package includes a comprehensive histopathology dataset with 250 cancer patients and 38 variables including demographics, pathological features, biomarkers, and outcomes.

```{r}
# Load the example dataset
data(histopathology)

# Examine the structure
str(histopathology, give.attr = FALSE)
```

```{r}
# Quick overview of key variables
histopathology %>%
  select(Age, Sex, Grade, TStage, Group, Outcome, OverallTime) %>%
  head(10)
```

## Basic Descriptive Analysis

### Table One

Generate a comprehensive descriptive summary table - the cornerstone of clinical research manuscripts:

```{r}
# Basic Table One
tableone(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI")
)
```

### Continuous Variables Summary

Generate detailed summaries for continuous variables:

```{r}
summarydata(
  data = histopathology,
  vars = c("Age", "OverallTime", "MeasurementA"),
  distr = TRUE,          # add normality / distribution diagnostics
  decimal_places = 2,    # digits used throughout the table
  outliers = TRUE,       # flag outlying values
  report_sentences = TRUE # copy-ready summary sentences
)
```

### Categorical Variables Summary

Analyze categorical variables with counts and percentages:

```{r}
# `vars` is restricted to categorical variables (permitted: factor), so use
# Grade_Level rather than the numeric Grade / TStage columns.
reportcat(
  data = histopathology,
  vars = c("Sex", "Grade_Level", "LVI", "PNI", "Group")
)
```

## Cross-tabulation Analysis

Create contingency tables with appropriate statistical tests:

```{r}
crosstable(
  data = histopathology,
  vars = c("Grade"),
  group = "Group"
)
```

```{r}
# Multiple variables cross-tabulation
crosstable(
  data = histopathology,
  vars = c("Grade", "TStage"),
  group = "Group"
)
```

## Data Visualization

### Age Pyramid

Visualize age and sex distribution:

```{r}
# `female` and `male` are Level options: jamovi makes them REQUIRED arguments
# of the R wrapper, so both must be supplied explicitly.
agepyramid(
  data = histopathology,
  age = "Age",
  gender = "Sex",
  female = "Female",
  male = "Male",
  bin_width = 5
)
```

### Alluvial Diagrams

Show relationships between categorical variables:

```{r}
alluvial(
  data = histopathology,
  vars = c("Grade", "TStage", "LVI"),
  condensationvar = "PNI"
)
```

### Venn Diagrams

Analyze overlap between sets:

```{r}
# var1..var7 must be categorical (permitted: factor), so the existing
# Present/Absent columns can be used directly - no recoding needed.
#
# var1true..var7true are Level options. jamovi forbids defaults on Level, so all
# seven are REQUIRED arguments of the R wrapper: pass NULL for the slots whose
# parent variable is not supplied.
venn(
  data = histopathology,
  var1 = "LVI",
  var1true = "Present",
  var2 = "PNI",
  var2true = "Present",
  var3 = "PreinvasiveComponent",
  var3true = "Present",
  var4 = "Grade_Level",
  var4true = "high",
  var5 = NULL,
  var5true = NULL,
  var6 = NULL,
  var6true = NULL,
  var7 = NULL,
  var7true = NULL
)
```

## Data Quality Assessment

### Benford's Law Analysis

Assess data quality using Benford's law - useful for detecting fabricated data:

```{r}
# Analyze first digits of age data (suitable for Benford's Law)
# benford(
#   data = histopathology,
#   var = "Age"
# )
```

### Variable Tree Summary

Get a hierarchical overview of your dataset:

```{r}
# Variable tree visualization.
# vartree has five Level options (percvarLevel, pruneLevel1, pruneLevel2,
# followLevel1, followLevel2). Level options can never carry a default, so all
# five are REQUIRED arguments of the R wrapper - pass NULL when the parent
# Variable (percvar / prunebelow / follow) is not supplied. The optional
# Variable arguments of vartree carry no default either, so pass them too.
vartree(
  data = histopathology,
  vars = c("Grade_Level", "LVI", "PNI"),  # vars is restricted to factors
  percvar = NULL,
  percvarLevel = NULL,
  summaryvar = NULL,
  prunebelow = NULL,
  pruneLevel1 = NULL,
  pruneLevel2 = NULL,
  follow = NULL,
  followLevel1 = NULL,
  followLevel2 = NULL,
  horizontal = TRUE,
  showcount = TRUE
)
```

## Working with Treatment Response Data

Load the treatment response dataset for oncology-specific analyses:

```{r}
# Load treatment response data
data(treatmentResponse)
head(treatmentResponse)
```

### Waterfall Plot

> **Not in this module.** `waterfall()` and `swimmerplot()` ship with the **OncoPath**
> module (`Analyses > OncoPath > Patient Follow-Up Plots`). The `treatmentResponse`
> dataset is bundled here for convenience, but the plotting analyses are not.

```{r}
# The bundled dataset has two columns: PatientID and ResponseValue
data(treatmentResponse)
head(treatmentResponse)
```

## Integration with jamovi

All functions in this package are also available through jamovi's graphical interface:

1. **Install the module** in jamovi
2. **Navigate to**: `Exploration > ClinicoPath Descriptives`
3. **Select your analysis** function
4. **Configure options** through the GUI
5. **Generate results** with point-and-click ease

### jamovi Advantages

- No programming required
- Interactive parameter adjustment
- Real-time result updates
- Export capabilities for reports
- Accessible to non-programmers

## Next Steps

This vignette covered the basic functionality of ClinicoPathDescriptives. For more advanced usage, see:

- **Clinical Workflow Vignette**: Complete analysis workflows for clinical research
- **Treatment Response Analysis**: Advanced oncology analysis techniques  
- **Data Quality and Exploration**: Comprehensive data assessment methods
- **Visualization Gallery**: Advanced plotting techniques and customization

## Citation

```{r eval=FALSE}
citation("ClinicoPathDescriptives")
```

## Support

- 📖 [Documentation](https://www.serdarbalci.com/ClinicoPathDescriptives/)
- 🐛 [Report Issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues)
- 🌐 [jamovi Module Hub](https://sbalci.github.io/ClinicoPathJamoviModule/)
