---
title: "ClinicoPath Descriptives 01: Introduction to Descriptive Statistics for Pathologists"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

The ClinicoPathDescriptives domain provides comprehensive tools for descriptive statistics specifically designed for clinical and pathological research. This module bridges the gap between traditional statistical analysis and specialized needs of medical researchers, offering clinically meaningful output formats and interpretations.

## What is ClinicoPathDescriptives?

ClinicoPathDescriptives is a specialized suite of statistical analysis tools within the ClinicoPath package that focuses on:

- **Clinical Data Summarization**: Generate Table One summaries, cross-tabulations, and descriptive statistics tailored for medical research
- **Quality Assessment**: Data quality checks, missing data analysis, outlier detection, and Benford's law screening
- **Visual Exploration**: Age pyramids, alluvial diagrams, Venn diagrams, and variable trees
- **Enhanced Reporting**: Medical terminology, clinical guidelines compliance, and journal-ready outputs

> **Scope note.** Agreement / inter-rater reliability analyses (`agreement()`, `icccoeff()`)
> are **not** part of ClinicoPathDescriptives - they are provided by the **meddecide**
> module. This module ships exactly fourteen analyses: `agepyramid`, `alluvial`, `benford`,
> `categorize`, `checkdata`, `chisqposttest`, `crosstable`, `dataquality`,
> `outlierdetection`, `reportcat`, `summarydata`, `tableone`, `vartree`, `venn`.

## Key Features

### 1. Table One Generation

`tableone()` takes a single `vars` list. There is no `explanatory`/`dependent` split -
use `crosstable()` when you need a grouping (column) variable.

```{r table-one-example}
# Generate comprehensive Table One
library(ClinicoPathDescriptives)
data(histopathology)

# Basic Table One
tableone(
  data = histopathology,
  vars = c("Grade", "Age", "MeasurementA"),
  sty = "t1",   # t1 = tableone, t2 = gtsummary, t3 = arsenal, t4 = janitor
  excl = FALSE
)
```

### 2. Advanced Cross-tabulations

`crosstable()` uses `vars` (the row variables) and `group` (the column variable).

```{r crosstable-example}
# Enhanced cross-tabulation with medical formatting
crosstable(
  data = histopathology,
  vars = c("Grade", "LVI"),
  group = "Group",  # must be a factor - `Outcome` is stored as numeric 0/1
  sty = "nejm",     # arsenal | finalfit | gtsummary | nejm | lancet | hmisc
  pcat = "chisq"    # chisq | fisher
)
```

### 3. Agreement Analysis

> **Not in this module.** `agreement()` and the `breast_agreement_data` dataset are
> provided by the **meddecide** module (`Analyses > meddecide > Agreement`), not by
> ClinicoPathDescriptives. The chunk below is kept for reference only and is not
> evaluated.

```{r agreement-example, eval=FALSE}
# Inter-rater agreement analysis - requires the meddecide module
# meddecide::agreement(...)
```

### 4. Data Quality Assessment

```{r data-quality-example}
# Comprehensive data quality check
dataquality(
  data = histopathology,
  vars = c("Age", "Grade", "TStage", "MeasurementA"),
  check_missing = TRUE,
  check_duplicates = TRUE,
  missing_threshold_visual = 10,  # percent, not a proportion
  plot_missing_patterns = TRUE,
  showSummary = TRUE,
  showRecommendations = TRUE
)
```

## Clinical Applications

### Pathology Research
- Digital pathology validation studies
- Biomarker concordance analysis
- Multi-platform validation
- Quality assurance metrics

### Clinical Trials
- Baseline characteristics tables
- Enrollment summaries
- Safety reporting
- Efficacy endpoints description

### Epidemiological Studies
- Population characteristics
- Risk factor analysis
- Screening program evaluation
- Public health surveillance

## Integration with Clinical Workflows

ClinicoPathDescriptives is designed to integrate seamlessly with:

1. **Laboratory Information Systems (LIS)**
2. **Electronic Health Records (EHR)**
3. **Research Electronic Data Capture (REDCap)**
4. **Clinical Data Management Systems**

## Available Functions

These are the fourteen analyses shipped by ClinicoPathDescriptives.

### Core Descriptive Functions
- `tableone()`: Comprehensive Table One generation (`sty`: t1/t2/t3/t4)
- `crosstable()`: Cross-tabulation with statistical tests (`vars` x `group`)
- `summarydata()`: Descriptive statistics for continuous variables
- `reportcat()`: Counts and percentages for categorical variables

### Data Quality Functions
- `dataquality()`: Multi-variable visual quality assessment
- `checkdata()`: Single-variable quality check (completeness, outliers, patterns)
- `outlierdetection()`: Univariate / multivariate / composite outlier identification
- `benford()`: Benford's law screening for fabricated or unusual numeric data
- `categorize()`: Convert a continuous variable into categories

### Visualization and Exploration
- `agepyramid()`: Age-sex pyramid
- `alluvial()`: Alluvial / Sankey diagrams
- `venn()`: Venn and UpSet diagrams (up to 7 sets)
- `vartree()`: Hierarchical variable tree
- `chisqposttest()`: Chi-square post-hoc pairwise comparisons

### Functions provided by other modules

Documentation occasionally refers to the following; they are **not** installed with
ClinicoPathDescriptives:

- `agreement()`, `icccoeff()` - **meddecide** module
- `waterfall()`, `swimmerplot()` - **OncoPath** module
- `groupsummary()`, `missingdata()`, `enhancedcrosstable()` - present in the upstream
  ClinicoPath development module only; not shipped here
- `gtsummary`, `tinytable` - third-party R packages used internally for table styling,
  not analyses in this module

## Getting Started

### Installation and Setup
```{r installation}
# ClinicoPath is available through jamovi
# For R users:
# install.packages("ClinicoPath") # When available on CRAN
# Or development version:
# remotes::install_github("sbalci/ClinicoPathJamoviModule")

library(ClinicoPathDescriptives)
```

### Basic Workflow
```{r basic-workflow}
# 1. Load your clinical data
data(histopathology)

# 2. Check data quality (`vars` is required)
quality_check <- dataquality(
  data = histopathology,
  vars = c("Age", "Grade", "MeasurementA"),
  check_missing = TRUE
)

# 3. Generate descriptive statistics
table_one <- tableone(
  data = histopathology,
  vars = c("Age", "Grade", "MeasurementA")
)

# 4. Create cross-tabulations
cross_tab <- crosstable(
  data = histopathology,
  vars = c("Grade"),
  group = "Group"
)

# 5. Export results
# Results are automatically formatted for clinical reporting
```

## Best Practices

### 1. Data Preparation
- Ensure consistent variable naming
- Use appropriate data types (factors for categorical variables)
- Handle missing data appropriately
- Validate data entry errors

### 2. Variable Selection
- Include clinically relevant variables
- Consider regulatory requirements
- Balance comprehensiveness with clarity
- Group related variables logically

### 3. Statistical Considerations
- Choose appropriate tests for data types
- Consider multiple testing corrections
- Report confidence intervals
- Include effect sizes when relevant

### 4. Reporting Standards
- Follow CONSORT guidelines for trials
- Adhere to STROBE guidelines for observational studies
- Use standard medical terminology
- Include appropriate disclaimers

## Related Vignettes

For more specific applications, see:

- [Digital Pathology Validation Studies](05-digital-pathology.html)
- [Modern Molecular Pathology Workflows](06-molecular-workflow.html)
- [Data Quality and Exploration](04-data-quality-exploration.html)
- [Comprehensive Data Quality Guide](01-comprehensive-data-quality.html)
- [Clinical Workflow](clinical-workflow.html)

## Support and Resources

### Documentation
- Function reference: `?tableone`, `?crosstable`, etc.
- Package documentation: `help(package = "ClinicoPathDescriptives")`
- Online documentation: [ClinicoPathDescriptives Website](https://www.serdarbalci.com/ClinicoPathDescriptives/)

### Community
- GitHub Issues: Report bugs and request features
- Discussion Forum: Share experiences and best practices
- Training Materials: Workshops and tutorials

### Citation
When using ClinicoPathDescriptives in your research, please cite:

```
Balci S (2024). ClinicoPath: Clinical and Pathological Data Analysis.
R package version X.X.X. https://github.com/sbalci/ClinicoPathJamoviModule
```

## Conclusion

ClinicoPathDescriptives provides a comprehensive suite of tools specifically designed for clinical and pathological research. By combining statistical rigor with clinical relevance, it enables researchers to generate publication-ready results while maintaining compliance with medical research standards.

The module's integration with jamovi makes advanced statistical analysis accessible to clinical researchers regardless of their programming background, while R users benefit from programmatic access and reproducible workflows.
