---
title: "Data Quality Assessment and Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Quality Assessment and Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Data quality assessment is fundamental to reliable clinical research. This vignette demonstrates essential data exploration and quality assessment techniques using ClinicoPathDescriptives.

## Basic Setup

```{r setup}
library(ClinicoPathDescriptives)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# Load the dataset
data(histopathology)

# Initial data overview
cat("Dataset Dimensions:", dim(histopathology), "\n")
cat("Variables:", ncol(histopathology), "\n")
cat("Observations:", nrow(histopathology), "\n\n")
```

## Phase 1: Completeness Assessment

### Missing Data Analysis

```{r}
# Calculate missing data percentages
missing_summary <- histopathology %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(
    Missing_Percentage = round(Missing_Count / nrow(histopathology) * 100, 1),
    Data_Quality = case_when(
      Missing_Percentage == 0 ~ "Complete",
      Missing_Percentage <= 5 ~ "Excellent", 
      Missing_Percentage <= 10 ~ "Good",
      Missing_Percentage <= 20 ~ "Acceptable",
      TRUE ~ "Poor"
    )
  ) %>%
  arrange(desc(Missing_Percentage))

# Display missing data summary
kable(missing_summary %>% filter(Missing_Count > 0),
      caption = "Missing Data Summary",
      col.names = c("Variable", "Missing Count", "Missing %", "Quality Rating"))
```

## Phase 2: Distribution Assessment

### Continuous Variables Analysis

```{r}
# Analyze distributions of continuous variables
continuous_vars <- c("Age", "OverallTime", "MeasurementA", "MeasurementB")

# Basic summary statistics
summary(histopathology[continuous_vars])

# Check for normality and distribution patterns
cat("\nDistribution Assessment:\n")
for(var in continuous_vars) {
  if(var %in% names(histopathology)) {
    cat(paste(var, "- Mean:", round(mean(histopathology[[var]], na.rm = TRUE), 2), 
              "SD:", round(sd(histopathology[[var]], na.rm = TRUE), 2), "\n"))
  }
}
```

### Categorical Variables Analysis

```{r}
# Analyze categorical variable distributions
categorical_vars <- c("Sex", "Grade", "TStage", "Group", "LVI", "PNI")

# Basic frequency tables for categorical variables
cat("Categorical Variables Distribution:\n\n")
for(var in categorical_vars) {
  if(var %in% names(histopathology)) {
    cat(paste("==", var, "==\n"))
    print(table(histopathology[[var]], useNA = "ifany"))
    cat("\n")
  }
}
```

## Phase 3: Quality Assessment with ClinicoPath Functions

### Cross-tabulation Analysis

```{r}
# Example of proper crosstable usage
crosstable(
  data = histopathology,
  vars = c("Grade", "TStage"),
  group = "Group"
)
```

### Data Summary

```{r}
# Example of proper summarydata usage.
# summarydata() takes only: vars, distr, decimal_places, outliers, report_sentences.
summarydata(
  data = histopathology,
  vars = c("Age", "OverallTime"),
  distr = TRUE,
  outliers = TRUE
)
```

### Categorical Reporting

```{r}
# Example of proper reportcat usage.
# `vars` only accepts categorical variables, so use Grade_Level (character)
# rather than Grade (numeric).
reportcat(
  data = histopathology,
  vars = c("Sex", "Grade_Level", "LVI", "PNI")
)
```

## Phase 4: Advanced Quality Assessment

### Benford's Law Analysis

```{r eval=FALSE}
# Fraud detection using Benford's law
# Create a dataset suitable for Benford analysis
set.seed(123)
benford_data <- data.frame(
  financial_amounts = round(rlnorm(300, 8, 2))
)

# Run Benford analysis
benford(
  data = benford_data,
  var = "financial_amounts"
)
```

### Agreement Analysis

> **Not in this module.** `agreement()` is provided by the **meddecide** module, not by
> ClinicoPathDescriptives. The chunk below is reference-only and is not evaluated.

```{r eval=FALSE}
# Requires the meddecide module to be installed
# meddecide::agreement(
#   data = histopathology,
#   vars = c("Rater 1", "Rater 2")
# )
```

## Best Practices Summary

### Data Quality Checklist

1. **Completeness**: Check for missing data patterns
2. **Consistency**: Verify logical relationships between variables
3. **Accuracy**: Assess distribution normality and outliers
4. **Validity**: Confirm data formats and ranges
5. **Reliability**: Test inter-rater agreement when applicable

### Quality Thresholds

- **Missing Data**: < 5% excellent, 5-10% good, 10-20% acceptable
- **Agreement**: κ > 0.8 excellent, 0.6-0.8 good, 0.4-0.6 moderate
- **Benford Compliance**: Chi-square p > 0.05 suggests natural data

## Conclusion

Regular data quality assessment using these techniques ensures reliable and trustworthy clinical research results. For detailed function usage and advanced examples, consult the individual function vignettes.

## Best Practices

- Perform quality assessment before any statistical analysis
- Document and investigate any quality issues found
- Establish quality thresholds appropriate for your research context
- Implement ongoing quality monitoring throughout data collection
- Use multiple quality assessment methods for comprehensive evaluation
