---
title: "Summarising Clinicopathological Data"
author: "ClinicoPathDescriptives Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising Clinicopathological Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette focuses on the functions that generate tables and textual summaries.

## Continuous Variables

`summarydata()` provides descriptive statistics (mean, median, quartiles) for numeric
variables. Beyond `vars` it accepts `distr`, `decimal_places`, `outliers` and
`report_sentences`.

```{r cont-sum, eval=FALSE}
summarydata(
  data = histopathology,
  vars = vars(Age, OverallTime),  # `vars` is restricted to numeric variables
  distr = TRUE,
  decimal_places = 2
)
```

## Categorical Variables

`reportcat()` summarises categorical variables with counts and percentages. Its `vars`
argument only accepts categorical variables, so use `Grade_Level` (character) rather than
`Grade` (numeric).

```{r cat-sum, eval=FALSE}
reportcat(
  data = histopathology,
  vars = vars(Sex, Grade_Level)
)
```

## Cross Tables

Use `crosstable()` to create cross tabulations with statistical tests. The `group`
argument defines the column variable and must be categorical. Styles (`sty`) are
`arsenal`, `finalfit`, `gtsummary`, `nejm` (default), `lancet` and `hmisc`.

```{r crosstable-vig, eval=FALSE}
crosstable(
  data = histopathology,
  vars = vars(Sex, Grade),
  group = "PreinvasiveComponent",
  sty = "nejm",
  excl = TRUE,
  cont = "mean",   # mean | median
  pcat = "chisq"   # chisq | fisher
)
```

## Table One for Publications

`tableone()` creates publication-ready tables of baseline characteristics. `sty` selects
the rendering engine: `t1` = tableone (default), `t2` = gtsummary, `t3` = arsenal,
`t4` = janitor.

```{r tableone-vig, eval=FALSE}
tableone(
  data = histopathology,
  vars = vars(Sex, PreinvasiveComponent, LVI, PNI, Grade, Age),
  sty = "t3",
  excl = TRUE
)
```

These functions output results objects that contain tables which can be converted to data frames for further customisation.

