---
title: "General 02: Getting Started with Statistical Analysis for Pathologists"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

ClinicoPathDescriptives is a comprehensive R package designed for descriptive analysis in clinicopathological research. It provides both programmatic R functions and a graphical user interface through jamovi, making advanced statistical analysis accessible to researchers with varying levels of programming experience.

### Key Features

- **Descriptive Statistics**: Generate comprehensive "Table One" summaries
- **Cross-tabulations**: Create contingency tables with appropriate statistical tests  
- **Visualization Tools**: Age pyramids, alluvial diagrams, Venn diagrams, and more
- **Data Quality Assessment**: Benford's law analysis, single- and multi-variable quality checks, outlier detection
- **Dual Interface**: Use in R or through jamovi's point-and-click interface

## Installation

### Install from GitHub

```{r eval=FALSE}
# Install development version
if (!require("remotes")) install.packages("remotes")
remotes::install_github("sbalci/ClinicoPathDescriptives")
```

### Install as jamovi Module

Download the `.jmo` file from [Releases](https://github.com/sbalci/ClinicoPathDescriptives/releases) and install in jamovi via: `jamovi menu > Modules > Install from file`

## Loading the Package

```{r setup}
library(ClinicoPathDescriptives)
library(dplyr)    # For data manipulation
library(ggplot2)  # For additional plotting
library(tidyr)   # For data reshaping 
```

## Example Dataset

The package includes a comprehensive histopathology dataset with 250 cancer patients and 38 variables including demographics, pathological features, biomarkers, and outcomes.

```{r}
# Load the example dataset
data(histopathology)

# Examine the structure
str(histopathology, give.attr = FALSE)
```

```{r}
# Quick overview of key variables
histopathology %>%
  select(Age, Sex, Grade, TStage, Group, Outcome, OverallTime) %>%
  head(10)
```

## Basic Descriptive Analysis

### Table One

Generate a comprehensive descriptive summary table - the cornerstone of clinical research manuscripts:

```{r}
# Basic Table One
tableone(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI")
)
```

### Continuous Variables Summary

Generate detailed summaries for continuous variables:

```{r}
summarydata(
  data = histopathology,
  vars = c("Age", "OverallTime", "MeasurementA"),
  distr = TRUE,          # add normality / distribution diagnostics
  decimal_places = 2,
  outliers = FALSE,
  report_sentences = FALSE
)
```

### Categorical Variables Summary

Analyze categorical variables with counts and percentages:

```{r}
reportcat(
  data = histopathology,
  vars = c("Sex", "Grade", "TStage", "LVI", "PNI")
)
```

## Cross-tabulation Analysis

Create contingency tables with appropriate statistical tests:

```{r}
crosstable(
  data = histopathology,
  vars = "Grade",
  group = "Group"
)
```

```{r}
# Multiple variables cross-tabulation
crosstable(
  data = histopathology,
  vars = c("Grade", "TStage"),
  group = "LymphNodeMetastasis"
)
```

## Data Visualization

### Age Pyramid

Visualize age and sex distribution. `female` and `male` are `Level` options, so the R
wrapper treats them as required arguments - always name both explicitly:

```{r}
agepyramid(
  data = histopathology,
  age = "Age",
  gender = "Sex",
  female = "Female",
  male = "Male",
  age_groups = "custom",   # custom | pediatric | reproductive | geriatric | lifecourse
  bin_width = 5
)
```

### Alluvial Diagrams

Show relationships between categorical variables:

```{r}
alluvial(
  data = histopathology,
  vars = c("Grade", "TStage", "LymphNodeMetastasis", "Outcome"),
  condensationvar = "Outcome"
)
```

### Venn Diagrams

Analyze overlap between sets:

```{r}
# Create binary indicators for Venn diagram
histopathology_venn <- histopathology %>%
  mutate(
    LVI_present = ifelse(LVI == "Present", 1, 0),
    PNI_present = ifelse(PNI == "Present", 1, 0),
    LN_positive = ifelse(LymphNodeMetastasis == "Present", 1, 0)
  )

# venn supports up to seven sets. var1true ... var7true are `Level` options and are
# therefore required arguments of the R wrapper - pass NULL for any unused set.
venn(
  data = histopathology_venn,
  var1 = "LVI_present",
  var1true = "1",
  var2 = "PNI_present",
  var2true = "1",
  var3 = "LN_positive",
  var3true = "1",
  var4 = NULL,
  var4true = NULL,
  var5 = NULL,
  var5true = NULL,
  var6 = NULL,
  var6true = NULL,
  var7 = NULL,
  var7true = NULL
)
```

## Data Quality Assessment

### Missing Value Analysis

Check for missing values and data completeness:

```{r}
# Check data completeness
missing_summary <- histopathology %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(variable, missing_count) %>%
  mutate(missing_percent = round(missing_count / nrow(histopathology) * 100, 1)) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))

print(missing_summary)
```

### Variable Tree Summary

Get a hierarchical overview of your dataset:

```{r eval=FALSE}
# Variable tree visualization example
# Note: vartree requires specific parameter configuration for optimal results
# This is a demonstration of the function call structure
vartree_result <- vartree(
  data = histopathology,
  vars = c("Grade", "TStage", "LVI", "PNI"),
  percvar = NULL,
  percvarLevel = NULL,
  summaryvar = NULL,
  prunebelow = NULL,
  pruneLevel1 = NULL,
  pruneLevel2 = NULL,
  follow = NULL,
  followLevel1 = NULL,
  followLevel2 = NULL,
  excl = FALSE,
  vp = TRUE,
  horizontal = FALSE,
  sline = TRUE,
  varnames = FALSE,
  nodelabel = TRUE,
  pct = FALSE
)

# For detailed vartree examples, see the vartree-specific vignette
cat("Variable tree visualization complete. See vartree documentation for advanced usage.")
```

## Working with Treatment Response Data

Load the treatment response dataset for oncology-specific analyses:

```{r}
# Load treatment response data
data(treatmentResponse)
if(exists("treatmentResponse")) {
  head(treatmentResponse)
} else {
  cat("treatmentResponse dataset not found. Using histopathology dataset instead.\n")
  head(histopathology)
}
```

### Waterfall Plot

> **Note:** `waterfall()` is **not** part of ClinicoPathDescriptives. Treatment-response
> visualisations (waterfall, swimmer, spider plots) are provided by the companion
> **OncoPath** module. The chunk below is kept for reference only and is not evaluated;
> install OncoPath to run it.

Visualize individual patient treatment responses:

```{r eval=FALSE}
# Waterfall plot example (requires proper treatmentResponse dataset)
if(exists("treatmentResponse") && "PatientID" %in% names(treatmentResponse)) {
  waterfall(
    data = treatmentResponse,
    patientID = "PatientID",
    responseVar = "ResponseValue",
    timeVar = NULL,
    showWaterfallPlot = TRUE
  )
} else {
  cat("Waterfall plot requires treatmentResponse dataset with PatientID and ResponseValue columns.\n")
  cat("See waterfall vignette for detailed examples with proper data structure.\n")
}
```

## Integration with jamovi

All functions in this package are also available through jamovi's graphical interface:

1. **Install the module** in jamovi
2. **Navigate to**: `Exploration > ClinicoPath Descriptives`
3. **Select your analysis** function
4. **Configure options** through the GUI
5. **Generate results** with point-and-click ease

### jamovi Advantages

- No programming required
- Interactive parameter adjustment
- Real-time result updates
- Export capabilities for reports
- Accessible to non-programmers

## Next Steps

This vignette covered the basic functionality of ClinicoPathDescriptives. For more advanced usage, see:

- **Clinical Workflow Vignette**: Complete analysis workflows for clinical research
- **Treatment Response Analysis**: Advanced oncology analysis techniques (provided by the separate **OncoPath** module)
- **Data Quality and Exploration**: Comprehensive data assessment methods
- **Visualization Gallery**: Advanced plotting techniques and customization

## Citation

```{r eval=FALSE}
citation("ClinicoPathDescriptives")
```

## Support

- 📖 [Documentation](https://www.serdarbalci.com/ClinicoPathDescriptives/)
- 🐛 [Report Issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues)
- 🌐 [jamovi Module Hub](https://sbalci.github.io/ClinicoPathJamoviModule/)
