---
title: "Vignette: Creating an Age Pyramid with ClinicoPathDescriptives"
author: "Analysis by Gemini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette: Creating an Age Pyramid with ClinicoPathDescriptives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# Create a sample data frame
set.seed(123) # for reproducibility
cohort_data <- data.frame(
  patient_id = 1:150,
  age = sample(20:85, 150, replace = TRUE),
  gender = sample(c("F", "M"), 150, replace = TRUE, prob = c(0.6, 0.4))
)

# View the first few rows
head(cohort_data)
```


```{r}
# This code simulates how the jamovi module would be called in an R environment.
# You would need the ClinicoPathDescriptives package installed.

# Load the library
library(ClinicoPathDescriptives)

# Example 1: Basic age pyramid analysis
# NOTE: `female` and `male` are `Level` options. jamovi forbids defaults on
# `Level`, so BOTH are required arguments of the R wrapper and must be passed
# explicitly (pass NULL if the corresponding level is not present in the data).
results1 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F", # Specify that 'F' represents females
    male = "M",   # Specify that 'M' represents males
    bin_width = 10, # Use 10-year age bins
    plot_title = "Age Distribution of Study Cohort"
)

# To view the results:
# View the summary table
print(results1$pyramidTable)

# View the plot
print(results1$plot)
```

## Example 2: Customizing Colors

Bar colours are chosen with `color_palette`, which accepts `"standard"` (the
default), `"colorblind"`, `"grayscale"` or `"custom"`. The `female_color` and
`male_color` parameters are only honoured when `color_palette = "custom"`:

```{r}
# Age pyramid with custom colors
results2 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F",
    male = "M",
    bin_width = 5, # Use 5-year age bins for more detail
    plot_title = "Age Distribution with Custom Colors",
    color_palette = "custom",  # Required for female_color / male_color to apply
    female_color = "#E74C3C",  # Red for the female bars
    male_color = "#3498DB"     # Blue for the male bars
)

# View the customized plot
print(results2$plot)
```

## Example 3: Different Age Bin Sizes

```{r}
# Compare different bin widths
# Fine granularity with 2-year bins
results3 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F",
    male = "M",
    bin_width = 2,
    plot_title = "Detailed Age Distribution (2-year bins)",
    color_palette = "custom",
    female_color = "#2ECC71", # Green
    male_color = "#F39C12"    # Orange
)

print(results3$plot)
```

## Example 4: Using Real Clinical Data

```{r}
# Load the histopathology dataset
data(histopathology)

# Create age pyramid for clinical cohort
clinical_results <- agepyramid(
    data = histopathology,
    age = "Age",
    gender = "Sex",
    female = "Female",
    male = "Male",
    bin_width = 10,
    plot_title = "Clinical Cohort Age Distribution",
    color_palette = "custom",
    female_color = "#9B59B6", # Purple
    male_color = "#1ABC9C"    # Teal
)

# View results
print(clinical_results$pyramidTable)
print(clinical_results$plot)
```

## Color Options

`color_palette` selects the colour scheme:

- `"standard"` (default) - the built-in pink/blue pair
- `"colorblind"` - a colour-vision-deficiency safe pair
- `"grayscale"` - print-friendly greys
- `"custom"` - use `female_color` and `male_color`

When `color_palette = "custom"`, `female_color` (default `"#E91E63"`) and
`male_color` (default `"#2196F3"`) accept any string R recognises as a colour:

- **Hex codes**: `"#FF5733"`, `"#3498DB"`
- **Named colors**: `"red"`, `"blue"`, `"darkgreen"`

## Age Grouping

`bin_width` (default `5`) sets equal-width age bins. Alternatively,
`age_groups` selects a preset grouping - `"custom"` (the default, which uses
`bin_width`, or `custom_breaks` when you supply a comma-separated list of cut
points), `"who"`, `"who_infant"`, `"pediatric"`, `"reproductive"`,
`"geriatric"` or `"lifecourse"`.

### WHO/UN standard age bands

Two presets give the published standard groupings:

- `"who"` - the WHO/UN standard five-year bands `0-4, 5-9, ... 80-84, 85+`
  (18 groups). These are the groups of the WHO World Standard Population
  (Ahmad OB *et al.*, *Age standardization of rates: a new WHO standard*,
  GPE Discussion Paper 31, WHO 2001) and the UN convention for population
  pyramids.
- `"who_infant"` - the WHO abridged life-table grouping, which reports infants
  separately from young children: `<1, 1-4, 5-9, ... 85+`.

```{r who-preset, eval=FALSE}
agepyramid(
    data       = histopathology,
    age        = "Age",
    gender     = "Sex",
    female     = "Female",
    male       = "Male",
    age_groups = "who"        # 0-4, 5-9, ... 85+
)
```

### Which bin edge is closed

`age_interval` controls how an age that falls exactly on a boundary is
assigned:

- `"left"` (the default) gives left-closed bands `[lower, upper)`. A person
  aged exactly 65 starts the `65-69` band, and equal-width bands hold equal
  numbers of single years. This is the WHO/UN and demographic convention, and
  it is what the presets above assume.
- `"right"` gives right-closed bands `(lower, upper]`, reproducing the
  behaviour of releases before 1.0.52. Note that this makes the lowest band one
  year wider than the others, which inflates the youngest bar of the pyramid.

Leave this at `"left"` unless you are reproducing an older analysis.

## Interpretation

The age pyramid plot shows:
- **Population distribution** by age groups and gender
- **Age structure** of your study cohort
- **Gender balance** across different age ranges
- **Data completeness** for demographic analysis

The accompanying table provides exact counts for each age group and gender category.

