---
title: "Vignette: Performing a Benford's Law Analysis"
author: "Analysis by Gemini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette: Performing a Benford's Law Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# Create a sample data frame of payment amounts
# These values are designed to roughly follow Benford's Law.
set.seed(123)
payments <- data.frame(
  amount = rlnorm(500, meanlog = 5, sdlog = 2)
)

# View the first few rows
head(payments)
```


## Running the analysis in ClinicoPathDescriptives

The `benford` analysis takes exactly two arguments beyond `data`:

- `var` - the continuous variable to test (numeric).
- `digits` - the number of leading digits to analyse, an integer between 1 and 3
  (default `2`). Values above 3 are not offered because `benford.analysis` does
  not provide a MAD conformity classification beyond three digits.

```{r}
library(ClinicoPathDescriptives)

# Default: two leading digits
results <- benford(
    data = payments,
    var = "amount",
    digits = 2
)

# First-digit only
results_1d <- benford(
    data = payments,
    var = "amount",
    digits = 1
)

# In jamovi the outputs render automatically. In R they are accessed as:
# results$summary          # conformity summary table
# results$text             # printed benford.analysis object
# results$text2            # suspect records
# results$plot             # digit-distribution plot
```

## What the module does under the hood

The chunk below reproduces the same computation with `benford.analysis`
directly, which is useful when you want to work with the fitted object.

```{r}
# Load required library
library(benford.analysis)

# Check the data structure
str(payments)
print(head(payments))

# Run the Benford analysis using the benford.analysis package directly
bfd.cp <- benford(data = payments$amount)

# View the results
print(bfd.cp)

# Create plot
plot(bfd.cp)

# Get suspects
suspects <- getSuspects(bfd.cp, payments)
if(length(suspects) > 0) {
  print("Suspicious values:")
  print(suspects)
} else {
  print("No suspicious values detected")
}
```

