---
title: "ClinicoPathDescriptives: Introduction to Clinical and Pathological Descriptive Statistics"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

The ClinicoPathDescriptives domain provides comprehensive tools for descriptive statistics specifically designed for clinical and pathological research. This module bridges the gap between traditional statistical analysis and specialized needs of medical researchers, offering clinically meaningful output formats and interpretations.

## What is ClinicoPathDescriptives?

ClinicoPathDescriptives is a specialized suite of statistical analysis tools within the ClinicoPath package that focuses on:

- **Clinical Data Summarization**: Generate Table One summaries, cross-tabulations, and descriptive statistics tailored for medical research
- **Quality Assessment**: Data quality checks, missing data analysis, outlier detection, and Benford's law screening
- **Distribution and Set Visualization**: Age pyramids, alluvial diagrams, variable trees, and Venn/UpSet plots
- **Enhanced Reporting**: Medical terminology, clinical guidelines compliance, and journal-ready outputs

Inter-rater agreement and diagnostic-test analyses are **not** part of
ClinicoPathDescriptives; they are provided by the companion **meddecide** module.

## Key Features

### 1. Table One Generation
```{r table-one-example}
# Generate comprehensive Table One
library(ClinicoPathDescriptives)
data(histopathology)

# Basic Table One
# `vars` takes every variable to describe; `sty` selects the table style
# (t1 = tableone, t2 = gtsummary, t3 = arsenal, t4 = janitor).
tableone(
  data = histopathology,
  vars = c("Grade", "Age", "TStage"),
  sty = "t1",
  excl = FALSE
)
```

### 2. Advanced Cross-tabulations
```{r crosstable-example}
# Enhanced cross-tabulation with medical formatting
# `vars` are the row variables, `group` is the column (grouping) variable.
crosstable(
  data = histopathology,
  vars = c("Grade", "TStage"),
  group = "Group",
  sty = "nejm",
  cont = "mean",
  pcat = "chisq"
)
```

### 3. Categorical and Continuous Summaries
```{r summary-example}
# Frequency summaries for categorical variables
reportcat(
  data = histopathology,
  vars = c("Grade", "LVI", "PNI")
)

# Descriptive statistics for continuous variables
summarydata(
  data = histopathology,
  vars = c("Age", "MeasurementA", "MeasurementB"),
  distr = TRUE,
  outliers = TRUE
)
```

### 4. Data Quality Assessment
```{r data-quality-example}
# Comprehensive data quality check across several variables
dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "MeasurementA"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  missing_threshold_visual = 10
)
```

## Clinical Applications

### Pathology Research
- Digital pathology validation studies
- Biomarker concordance analysis
- Multi-platform validation
- Quality assurance metrics

### Clinical Trials
- Baseline characteristics tables
- Enrollment summaries
- Safety reporting
- Efficacy endpoints description

### Epidemiological Studies
- Population characteristics
- Risk factor analysis
- Screening program evaluation
- Public health surveillance

## Integration with Clinical Workflows

ClinicoPathDescriptives is designed to integrate seamlessly with:

1. **Laboratory Information Systems (LIS)**
2. **Electronic Health Records (EHR)**
3. **Research Electronic Data Capture (REDCap)**
4. **Clinical Data Management Systems**

## Available Functions

ClinicoPathDescriptives ships exactly fourteen analyses.

### Core Descriptive Functions
- `tableone()`: Table One generation (styles `t1`-`t4`)
- `crosstable()`: Cross-tabulation with automatic test selection
- `summarydata()`: Descriptive statistics for continuous variables
- `reportcat()`: Frequency summaries for categorical variables
- `chisqposttest()`: Chi-square post-hoc pairwise comparisons

### Data Quality Functions
- `dataquality()`: Multi-variable visual quality assessment
- `checkdata()`: Single-variable quality check
- `outlierdetection()`: Outlier identification (univariate/multivariate/composite)
- `benford()`: Benford's law screening for data integrity
- `categorize()`: Convert continuous variables into categories

### Visualization Functions
- `agepyramid()`: Age and sex distribution pyramids
- `alluvial()`: Alluvial / Sankey flow diagrams
- `vartree()`: Hierarchical variable trees
- `venn()`: Venn diagrams and UpSet plots

### Functions provided by other ClinicoPath modules
The following are frequently confused with this module but live elsewhere:
`agreement()` and `icccoeff()` (**meddecide**), survival analyses
(**jsurvival**), `waterfall()` and `swimmerplot()` (**OncoPath**), and the
ggstatsplot wrappers (**jjstatsplot**).

## Getting Started

### Installation and Setup
```{r installation}
# ClinicoPath is available through jamovi
# For R users:
# install.packages("ClinicoPath") # When available on CRAN
# Or development version:
# remotes::install_github("sbalci/ClinicoPathJamoviModule")

library(ClinicoPathDescriptives)
```

### Basic Workflow
```{r basic-workflow}
# 1. Load your clinical data
data(histopathology)

# 2. Check data quality (`vars` is required)
quality_check <- dataquality(
  data = histopathology,
  vars = c("Age", "Grade", "TStage", "MeasurementA")
)

# 3. Generate descriptive statistics
table_one <- tableone(
  data = histopathology,
  vars = c("Age", "Grade", "TStage")
)

# 4. Create cross-tabulations
cross_tab <- crosstable(
  data = histopathology,
  vars = c("Grade", "TStage"),
  group = "Group"
)

# 5. Export results
# Results are automatically formatted for clinical reporting
```

## Best Practices

### 1. Data Preparation
- Ensure consistent variable naming
- Use appropriate data types (factors for categorical variables)
- Handle missing data appropriately
- Validate data entry errors

### 2. Variable Selection
- Include clinically relevant variables
- Consider regulatory requirements
- Balance comprehensiveness with clarity
- Group related variables logically

### 3. Statistical Considerations
- Choose appropriate tests for data types
- Consider multiple testing corrections
- Report confidence intervals
- Include effect sizes when relevant

### 4. Reporting Standards
- Follow CONSORT guidelines for trials
- Adhere to STROBE guidelines for observational studies
- Use standard medical terminology
- Include appropriate disclaimers

## Related Vignettes

For more specific applications, see:

- [Data Quality and Exploration](04-data-quality-exploration.html)
- [Digital Pathology](05-digital-pathology.html)
- [Molecular Pathology Workflow](06-molecular-workflow.html)
- [Digital Pathology Validation Studies](18-digital-pathology-validation-legacy.html)
- [Modern Molecular Pathology Workflows](19-molecular-pathology-workflow-legacy.html)

## Support and Resources

### Documentation
- Function reference: `?tableone`, `?crosstable`, etc.
- Package documentation: `help(package = "ClinicoPathDescriptives")`
- Online documentation: [ClinicoPathDescriptives Website](https://www.serdarbalci.com/ClinicoPathDescriptives/)

### Community
- GitHub Issues: Report bugs and request features
- Discussion Forum: Share experiences and best practices
- Training Materials: Workshops and tutorials

### Citation
When using ClinicoPathDescriptives in your research, please cite:

```
Balci S (2024). ClinicoPath: Clinical and Pathological Data Analysis.
R package version X.X.X. https://github.com/sbalci/ClinicoPathJamoviModule
```

## Conclusion

ClinicoPathDescriptives provides a comprehensive suite of tools specifically designed for clinical and pathological research. By combining statistical rigor with clinical relevance, it enables researchers to generate publication-ready results while maintaining compliance with medical research standards.

The module's integration with jamovi makes advanced statistical analysis accessible to clinical researchers regardless of their programming background, while R users benefit from programmatic access and reproducible workflows.
