---
title: "Use ClinicoPathDescriptives in R"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPathDescriptives)
```




```{r data histopathology}
data("histopathology")
```

```{r}
#| glimpse histopathology

dplyr::glimpse(histopathology)
```



`female` and `male` are `Level` options. The jamovi compiler forbids defaults on
`Level`, so both are **required** arguments of the generated R wrapper and must
always be supplied.

```{r}
ClinicoPathDescriptives::agepyramid(
    data = histopathology,
    age = Age,
    gender = Sex,
    female = "Female",
    male = "Male")
```



```{r}
age_pyramid <-
ClinicoPathDescriptives::agepyramid(
    data = histopathology,
    age = Age,
    gender = Sex,
    female = "Female",
    male = "Male")
```



```{r}
age_pyramid$pyramidTable
```


```{r}
age_pyramid$pyramidTable$asDF
```


```{r}
age_pyramid$plot
```



```{r}
age_pyramid$plot$state
```



```{r}
age_pyramid$plot$plot
```


```{r}
temp_file_png <- tempfile(fileext = ".png")
age_pyramid$plot$saveAs(paste0(temp_file_png, ".png"))

age_pyramid_plot <- png::readPNG(paste0(temp_file_png, ".png"))

temp_file_tiff <- tempfile(fileext = ".tiff")

tiff::writeTIFF(what = age_pyramid_plot,
                where = paste0(temp_file_tiff, ".tiff"),
                compression = "LZW")
```




```{r}
htmlTable::htmlTable(
  kableExtra::kable(
    age_pyramid$pyramidTable$asDF
    ))
```



## Analyses shipped by ClinicoPathDescriptives

Each analysis exposes a plain R function plus its R6 backend class
(`<name>Class`).

| Function | Title |
|---|---|
| `agepyramid()` | Age Pyramid |
| `alluvial()` | Alluvial Diagrams |
| `benford()` | Benford Analysis |
| `categorize()` | Categorize Continuous Variables |
| `checkdata()` | Single Variable Quality Check |
| `chisqposttest()` | Chi-Square Post-Hoc Tests |
| `crosstable()` | Cross Tables |
| `dataquality()` | Multi-Variable Visual Quality |
| `outlierdetection()` | Outlier Detection |
| `reportcat()` | Summary of Categorical Variables |
| `summarydata()` | Summary of Continuous Variables |
| `tableone()` | Table One |
| `vartree()` | Variable Tree |
| `venn()` | Venn Diagram |

The bundled example dataset used above is `histopathology`.

### Arguments that are always required

Options of type `Level` cannot carry a default, so they are always required
arguments of the generated R wrapper. Pass `NULL` when the parent variable is
not supplied:

- `agepyramid()`: `female`, `male`
- `venn()`: `var1true` ... `var7true`
- `vartree()`: `percvarLevel`, `pruneLevel1`, `pruneLevel2`, `followLevel1`,
  `followLevel2` (and the `Variable` options `vars`, `percvar`, `summaryvar`,
  `prunebelow`, `follow`, which likewise have no defaults)
