---
title: "Vignette: Visualizing Co-occurrence with Venn & Upset Plots"
author: "Analysis by Gemini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette: Visualizing Co-occurrence with Venn & Upset Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

```{r setup}
# For vignette demonstration, ensure 'histopathology' is available.
if (!exists("histopathology")) {
  set.seed(123)
  histopathology <- data.frame(
    LVI = factor(sample(c("Present", "Absent", NA), 150, replace = TRUE, prob = c(0.3, 0.6, 0.1))),
    PNI = factor(sample(c("Present", "Absent"), 150, replace = TRUE, prob = c(0.2, 0.8))),
    PreinvasiveComponent = factor(sample(c("Yes", "No"), 150, replace = TRUE, prob = c(0.4, 0.6)))
  )
}
head(histopathology)
```


`venn()` accepts up to seven variables. `var1true` ... `var7true` are `Level`
options; the jamovi compiler forbids defaults on `Level`, so all seven are
**required** arguments of the R wrapper and must be passed explicitly - `NULL`
for any variable slot left unused.

```{r}
library(ClinicoPathDescriptives)
# data(histopathology, package="ClinicoPathDescriptives") # if needed

results_venn <- ClinicoPathDescriptives::venn(
  data = histopathology,
  var1 = "LVI",
  var1true = "Present",
  var2 = "PNI",
  var2true = "Present",
  var3 = "PreinvasiveComponent",
  var3true = "Yes",
  var4 = NULL,      # Explicitly NULL for unused variable
  var4true = NULL,  # Explicitly NULL for unused true level
  var5 = NULL,
  var5true = NULL,
  var6 = NULL,
  var6true = NULL,
  var7 = NULL,
  var7true = NULL,
  show_ggvenn = TRUE,
  show_ggVennDiagram = TRUE
)

```

The plot outputs are named after the package that draws them: `plotGgvenn`
(shown by default via `show_ggvenn = TRUE`), `plotGgVennDiagram`, `plotUpsetR`
and `plotComplexUpset`. Each is rendered only when its corresponding
`show_*` option is `TRUE`.

```{r}
results_venn$plotGgvenn
```



```{r}
results_venn$plotGgVennDiagram
```


