---
title: "Vignette: Categorizing Continuous Variables for Clinicopathological Reporting"
author: "ClinicoPathDescriptives"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette: Categorize Continuous Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## What this analysis is for

`categorize()` turns one continuous numeric variable into a factor, using a
binning rule you choose, and reports the resulting break points, the frequency
of each category, a distribution plot with the cut-points drawn on it, and the
plain R code that reproduces the same recoding outside jamovi.

Pathologists do this constantly, usually by hand:

* Ki-67 labelling index into low / intermediate / high proliferation bands;
* ER / PR percentage into negative, low-positive, positive;
* eGFR into CKD stages at the conventional 30 / 60 / 90 thresholds;
* tumour size into pT-like strata;
* any laboratory analyte split at a reference cut-point before it goes into a
  cross-table or a Kaplan-Meier plot.

The point of the analysis is that the recoding is *visible*: you can see where
the boundaries landed, how many cases fell into each band, and whether the split
you chose produced a group of three patients that will not support any
subsequent test.

## The statistical caveat, up front

Cutting a continuous measurement into categories throws information away. It
costs power, it makes the result depend on where you happened to put the
boundary, and it hides dose-response relationships by pretending that two
patients on either side of a threshold differ while two patients at opposite
ends of the same band do not. Optimal or "data-derived" cut-points found by
trying several thresholds and keeping the most significant one are the worst
case: they inflate the type I error rate and rarely replicate.

> Altman DG, Royston P. **The cost of dichotomising continuous variables.**
> *BMJ* 2006;332:1080. doi:10.1136/bmj.332.7549.1080

The analysis itself repeats this citation in its completion notice. Categorize
when there is a clinical reason to - an established reporting threshold, a
guideline stage definition, a decision rule that a clinician will actually act
on - and keep the continuous variable in the model whenever you can.

## Example data

```{r}
set.seed(42)
n <- 240

path_data <- data.frame(
  Ki67_Index   = pmin(pmax(rnorm(n, mean = 22, sd = 15), 0), 95),
  TumorSize_mm = rlnorm(n, meanlog = 3, sdlog = 0.5),
  eGFR         = rnorm(n, mean = 68, sd = 22)
)

head(path_data)
```

```{r}
library(ClinicoPathDescriptives)
```

## Binning methods

The `method` option is a list with exactly six levels:
`equal`, `quantile` (the default), `manual`, `meansd`, `median`, `jenks`.

### `quantile` - equal numbers per group (default)

Break points are the sample quantiles at `seq(0, 1, length.out = nbins + 1)`,
de-duplicated. Every category holds roughly the same number of patients, which
is what you want for tertiles, quartiles or a median-free "balanced" split.

```{r}
res_q <- categorize(
    data   = path_data,
    var    = "Ki67_Index",
    method = "quantile",   # default
    nbins  = 4             # default 4 -> quartiles
)

res_q$freqTable
res_q$breakpointsTable
```

Tied values collapse bins: if you ask for four quartiles of a variable where a
quarter of the cases share one value, you may get three. The analysis raises a
*bin collapse* warning when the number of categories it managed to build differs
from `nbins`, because calling the result "quartiles" would then be wrong.

### `equal` - equal-width intervals

`seq(min(x), max(x), length.out = nbins + 1)`. Bands are the same width on the
measurement scale, so counts can be wildly unequal on a skewed variable such as
tumour size. Use it when the width itself is the meaningful thing (10 mm steps),
not when you want comparable group sizes.

```{r}
res_eq <- categorize(
    data   = path_data,
    var    = "TumorSize_mm",
    method = "equal",
    nbins  = 5
)
```

### `manual` - your own cut-points

`breaks` is a **string** of comma-separated numbers (it is a `String` option,
not a numeric vector). They are parsed, trimmed and sorted for you.

```{r}
res_ckd <- categorize(
    data   = path_data,
    var    = "eGFR",
    method = "manual",
    breaks = "30, 60, 90",     # the conventional CKD thresholds
    labels = "custom",
    customlabels = "Stage 4-5, Stage 3, Stage 2, Stage 1"
)

res_ckd$breakpointsTable
```

By default the outer break points are **extended** to the data minimum and
maximum - a value below 30 or above 90 is not lost, it simply joins the first or
last band. That is why the example above yields four categories from three
entered break points: the observed minimum and maximum are prepended and
appended. Count the categories before you write the labels.

`nbins` is ignored for this method; the number of categories is determined by
the break points.

### `meansd` - mean +/- k standard deviations

Break points are `c(min(x), mean - sdmult * sd, mean, mean + sdmult * sd,
max(x))`, i.e. **four** categories regardless of `nbins`. `sdmult` is a number
between 0.5 and 3, default 1.

```{r}
res_sd <- categorize(
    data   = path_data,
    var    = "Ki67_Index",
    method = "meansd",
    sdmult = 1        # default; 0.5 to 3 permitted
)
```

Both the mean and the SD are pulled by outliers, so a handful of extreme values
will drag every boundary. The analysis screens for extreme values with a 3 x IQR
rule and warns you when it finds any, suggesting `quantile` or `jenks` instead.
If the variable has zero variance the break points cannot be built and you get an
explicit validation error rather than a silent result.

### `median` - dichotomisation at the median

Break points are `c(min(x), median(x), max(x))`: two groups, `nbins` ignored.
Convenient, ubiquitous in the pathology literature, and precisely the practice
Altman and Royston were writing against. Prefer a guideline threshold via
`manual` when one exists.

```{r}
res_med <- categorize(
    data   = path_data,
    var    = "Ki67_Index",
    method = "median",
    labels = "semantic"    # -> "Low", "High"
)
```

### `jenks` - natural breaks

Fisher-Jenks natural breaks, computed with
`classInt::classIntervals(x, n = nbins, style = "jenks")`. It places boundaries
so that within-class variance is minimised, i.e. it cuts where the data are
already sparse. Useful for a genuinely multimodal marker.

`classInt` is consulted at run time; if it is not installed the analysis falls
back to quantile break points and tells you so in a warning. Install it with
`install.packages("classInt")` to get true Jenks optimisation.

```{r}
res_jk <- categorize(
    data   = path_data,
    var    = "Ki67_Index",
    method = "jenks",
    nbins  = 3
)
```

## Number of categories

`nbins` is an integer between 2 and 20, default 4. It is used by `equal`,
`quantile` and `jenks` only - `manual` takes its category count from the break
points, `meansd` always produces four bands and `median` always produces two.

## Labelling the categories

`labels` has five levels:

| `labels` | What you get |
|---|---|
| `auto` (default) | Interval notation built from the break points, e.g. `[0.0, 12.4]`, `(12.4, 20.7]` - brackets follow the closure options below |
| `semantic` | 2 bins: Low / High. 3: Low / Medium / High. 4: Low / Medium-Low / Medium-High / High. 5: Very Low / Low / Medium / High / Very High. 6-7: the first *n* of Very Low, Low, Medium-Low, Medium, Medium-High, High, Very High. More than 7: `Level 1`, `Level 2`, ... |
| `numbered` | `1`, `2`, `3`, ... |
| `lettered` | `A`, `B`, `C`, ... |
| `custom` | Your own names, from `customlabels` |

`customlabels` is a comma-separated string. It must supply exactly as many
labels as there are categories, and they must be unique; if the count does not
match or a label is duplicated the analysis warns and falls back to
`Category 1`, `Category 2`, ... rather than silently merging levels. Count the
categories first - remember that `manual` extends the outer breaks.

```{r}
res_lab <- categorize(
    data         = path_data,
    var          = "Ki67_Index",
    method       = "quantile",
    nbins        = 3,
    labels       = "custom",
    customlabels = "Low proliferation, Intermediate, High proliferation"
)
```

`newvarname` (default `""`) names the derived variable. Left empty, the name is
the original variable plus `_cat`, e.g. `Ki67_Index_cat`.

## Interval closure

These two options decide which side of a boundary a patient sitting exactly on
it falls, and they matter a great deal at a reporting threshold.

* `rightclosed` (default `TRUE`) - intervals are right-closed / left-open:
  `(a, b]`. A Ki-67 of exactly 20 with a break at 20 falls in the *lower* band.
  Set it to `FALSE` for left-closed / right-open intervals `[a, b)`, where 20
  falls in the *upper* band. This is the usual convention for "Ki-67 >= 20%
  is high", so set `rightclosed = FALSE` when your threshold is written with
  `>=`.
* `includelowest` (default `TRUE`) - the extreme interval is closed at the outer
  end so the minimum (with `rightclosed = TRUE`) or the maximum (with
  `rightclosed = FALSE`) is not left uncategorised.

Both map directly onto `cut()`'s `right =` and `include.lowest =` arguments, and
the bracket notation in the *Range* column of the frequency table follows
whatever you chose.

```{r}
# "Ki-67 >= 20% is high" - the threshold belongs to the upper band
res_ge20 <- categorize(
    data        = path_data,
    var         = "Ki67_Index",
    method      = "manual",
    breaks      = "20",
    rightclosed = FALSE,      # [a, b) -> exactly 20 counts as high
    includelowest = TRUE,
    labels      = "semantic"
)
```

`ordered` (default `TRUE`) makes the result an ordered factor, which preserves
Low < Medium < High in every downstream table and plot. Turn it off only if a
later analysis insists on an unordered factor.

`excl` (default `TRUE`) excludes missing values when the break points are
computed. Missing cases are still reported: the frequency table adds a
*Missing* row, and percentages are calculated over valid observations only.

## Outputs

| Output | Contents |
|---|---|
| `todo` | Instructions before a variable is chosen; afterwards, the validation errors and the warning / information notices |
| `summaryText` | N total, N valid, N missing, range, mean, median, SD of the input variable |
| `breakpointsTable` | Every break point, in order, rounded to 4 decimals |
| `freqTable` | Category, range, count, percent, cumulative percent (plus a Missing row when needed) |
| `plot` | Histogram with density curve and the break points as red dashed lines, shown when `showplot = TRUE` (default) |
| `rcode` | Copy-paste R that reproduces the recoding, shown when `showcode = TRUE` (default) |

```{r}
res <- categorize(
    data     = path_data,
    var      = "Ki67_Index",
    method   = "quantile",
    nbins    = 3,
    labels   = "semantic",
    showcode = TRUE,   # default
    showplot = TRUE    # default
)

res$summaryText
res$freqTable
res$breakpointsTable
res$rcode
```

## Warnings the analysis will raise

These are quality gates worth reading rather than dismissing:

* **Small bins** - at least one category has fewer than 5 observations. Any test
  built on that category will be unreliable.
* **Severe bin imbalance** - one category holds more than 70% of the valid
  observations, which usually means the cut-points are in the wrong place.
* **Bin collapse** - fewer distinct categories than `nbins` were achievable
  because of ties or limited range, so "quartiles" or "tertiles" is no longer an
  accurate description.
* **Custom labels mismatch / duplicates** - label count does not match the
  category count, or labels repeat; numbered labels are used instead.
* **Jenks fallback** - `classInt` is not installed and quantile breaks were used.
* **Outlier sensitivity** - extreme values detected while using `meansd`.
* **Completion note** - N, number of groups, method, and the Altman & Royston
  caveat.

## Adding the result back to the dataset

In jamovi, tick **Categorized variable** in the options panel (the `addtodata`
output). The new column appears in the spreadsheet immediately, named by
`newvarname` or `<variable>_cat`, and can be used straight away in Cross Tables,
Alluvial Diagrams, Table One, survival analyses and so on.

`addtodata` is an `Output`-type option: it exists to write a column into a
jamovi data set, and it is **not** an argument of the `categorize()` R function.
When you are working in R, take the generated code from the `rcode` output
instead - it is exactly the `cut()` call the analysis performed, and it appends
the new column to your own data frame:

```{r}
# The shape of what the rcode output prints for a quantile split:
breaks <- unique(sort(quantile(path_data$Ki67_Index,
                              probs = seq(0, 1, length.out = 4),
                              na.rm = TRUE)))
labels <- c("Low", "Medium", "High")

path_data$Ki67_Index_cat <- cut(
  path_data$Ki67_Index,
  breaks = breaks,
  labels = labels,
  include.lowest = TRUE,
  right = TRUE,
  ordered_result = TRUE
)

table(path_data$Ki67_Index_cat)
```

## Values outside your break points

With `method = "manual"`, your break points are used exactly as entered and the
range is then **extended** to the data minimum and maximum, so every observation
lands in a category and no case is lost. Entering `30,60,90` on eGFR data
spanning 12.3-109.2 therefore gives five break points - `12.26, 30, 60, 90,
109.21` - and four bins, which is what CKD staging wants.

That also means you get a bin below your lowest threshold and one above your
highest. If you would rather **drop** those observations and keep only the range
you specified, set `excludeoutofrange = TRUE`:

```{r exclude-oor, eval=FALSE}
categorize(
    data              = ckd_data,
    var               = "eGFR",
    method            = "manual",
    breaks            = "30,60,90",
    excludeoutofrange = TRUE      # keep 30-90 only; drop everything outside
)
```

The break points are then exactly `30, 60, 90` (two bins), and observations
below 30 or above 90 are excluded. The analysis reports how many went, split by
direction - for example *"Excluded 60 observation(s) (33.3%) that fall outside
the break points [30, 90]: 30 below and 30 above"* - so the loss is never
silent.

The option applies to `manual` only. The computed methods build their break
points from the data minimum and maximum, so nothing can fall outside them.

## Practical checklist

1. Have a reason to categorise that is clinical, not statistical.
2. Prefer an established threshold (`manual`) over a data-derived one.
3. Check `rightclosed` against how your threshold is written (`>` vs `>=`).
4. Read the break points table - especially after `manual`, where the outer
   breaks are extended.
5. Read the frequency table for tiny or dominant categories before you run
   anything downstream.
6. Report the cut-points and the rule in your methods section, and keep the
   continuous analysis alongside the categorical one.
