---
title: "ClinicoPath Descriptives: Oncology Data Examples"
author: "ClinicoPath Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ClinicoPath Descriptives: Oncology Data Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# OncoDataSets is an optional demo dependency and is NOT imported by this module.
# Gate all code execution on its availability so the site still builds when it is
# absent (code is shown for reference but not evaluated); runs fully if installed.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("OncoDataSets", quietly = TRUE)
)
```

> ### Scope note
>
> This vignette illustrates **jamovi menu workflows** for analyses that
> ClinicoPathDescriptives 1.0.52 really does ship — `tableone`, `crosstable`, `summarydata`,
> `checkdata` and `chisqposttest`. It calls no ClinicoPath R function directly; the R chunks
> only prepare `OncoDataSets` data, and the workflow chunks are commented menu instructions
> kept at `eval = FALSE`.
>
> The menu paths and option names below were checked against the shipped
> `jamovi/*.a.yaml` definitions of version 1.0.52. Every analysis in this module lives under
> the top-level **Exploration** menu.

## Introduction

This vignette demonstrates the use of ClinicoPath Descriptives module with real oncology datasets from the `OncoDataSets` package. We'll explore descriptive statistics, cross-tabulations, and data summaries using cancer research data.

## Loading Required Packages

```{r packages, message=FALSE}
library(ClinicoPathDescriptives)
library(OncoDataSets)
library(dplyr)
library(knitr)
```

## Example 1: Breast Cancer Wisconsin Dataset

The Wisconsin Breast Cancer dataset contains features computed from digitized images of fine needle aspirates (FNA) of breast masses.

```{r breast-cancer-data}
data("BreastCancerWI_df")

# Overview of the dataset
dim(BreastCancerWI_df)
names(BreastCancerWI_df)[1:10]  # First 10 variables

# Convert diagnosis to factor with meaningful labels
BreastCancerWI_df$diagnosis <- factor(BreastCancerWI_df$diagnosis, 
                                      levels = c("0", "1"),
                                      labels = c("Benign", "Malignant"))
```

### Table One Analysis

Create a baseline characteristics table for the cohort:

```{r table-one, eval=FALSE}
# In jamovi, you would:
# 1. Select Analyses → Exploration → ClinicoPath Descriptives → Table One
# 2. Add variables: radius_mean, texture_mean, perimeter_mean, area_mean, diagnosis
# 3. Choose a "Table style": t1 (tableone, default), t2 (gtsummary),
#    t3 (arsenal) or t4 (janitor)
# 4. Optionally tick "Missing-value exclusion (NA)", "Analysis summary",
#    "About this analysis" or "Copy-ready report text"
#
# Note: Table One in this module is UNGROUPED - it has no grouping/strata option.
# To compare benign vs malignant side by side, use Cross Tables (next section),
# which takes a grouping variable and reports a p-value per row.
```

### Cross-tabulation Example

Let's categorize tumor size and examine its relationship with diagnosis:

```{r cross-tab-prep}
# Categorize radius_mean
BreastCancerWI_df$radius_category <- cut(BreastCancerWI_df$radius_mean,
                                         breaks = c(0, 10, 15, 20, Inf),
                                         labels = c("Small", "Medium", "Large", "Very Large"))

# Summary
table(BreastCancerWI_df$radius_category, BreastCancerWI_df$diagnosis)
```

```{r cross-tab, eval=FALSE}
# In jamovi:
# 1. Select Analyses → Exploration → ClinicoPath Comparisons → Cross Tables
# 2. Put 'radius_category' in "Variables" and 'diagnosis' in "Group Variable"
#    (rows are the variables; the group variable forms the columns)
# 3. Pick the categorical test with "pcat": chisq (default) or fisher
# 4. Optional: "cont" = mean (default) or median for continuous rows,
#    "p_adjust" = none (default) / bonferroni / holm / BH / BY, and "showSMD"
#    for standardized mean differences
#
# Note: Cross Tables does not report standardized residuals. For residual analysis
# use Chi-Square Post-Hoc Tests (see the last section of this vignette).
```

## Example 2: Childhood Cancer Dataset

The childhood cancer dataset provides information about cancer cases in children.

```{r child-cancer}
data("ChildCancer_df")

# Dataset structure
str(ChildCancer_df)

# Summary statistics
summary(ChildCancer_df)
```

### Descriptive Analysis by Cancer Type

```{r child-cancer-desc, eval=FALSE}
# In jamovi:
# 1. Select Analyses → Exploration → ClinicoPath Descriptives →
#    Summary of Continuous Variables
# 2. Add the continuous variables (e.g. SurvTime) to "Variables"
# 3. Tick "Distribution diagnostics" (distr) for skewness/kurtosis and
#    normality checks; tick "Outlier detection" (outliers) to flag extreme values
# 4. Set "Decimal places" (default 2) and tick "Report sentences" for
#    copy-ready narrative text
#
# Note: this analysis has no grouping option. For categorical variables such as
# ICCGroup or Sex, use Summary of Categorical Variables instead; to describe one
# variable across the levels of another, use Cross Tables.
```

## Example 3: Bladder Cancer Dataset

```{r bladder-cancer}
data("BladderCancer_df")

# Examine the structure
head(BladderCancer_df)
str(BladderCancer_df)
```

### Data Quality Check

```{r data-check, eval=FALSE}
# In jamovi, for ONE variable at a time:
# 1. Select Analyses → Exploration → ClinicoPath Data Quality →
#    Single Variable Quality Check
# 2. Put a single variable (e.g. 'size') in "Variable" - this analysis accepts one
# 3. "Outlier analysis" (showOutliers) is on by default; optional extras are
#    "Distribution analysis", "Duplicate analysis", "Data patterns"
#    and "Little's MCAR test" (mcarTest)
#
# For several variables at once:
# 1. Select Analyses → Exploration → ClinicoPath Data Quality →
#    Multi-Variable Visual Quality
# 2. Add all variables, then tick "check_missing", "check_duplicates" and the
#    plot options ("plot_data_overview", "plot_missing_patterns", "plot_data_types")
#
# For pure outlier work across several numeric variables, use
# Exploration → ClinicoPath Data Quality → Outlier Detection.
```

## Example 4: Smoking and Lung Cancer Relationship

```{r smoking-lung}
data("SmokingLungCancer_df")

# Dataset overview
str(SmokingLungCancer_df)
head(SmokingLungCancer_df)
```

### Cross-tabulation of Aggregated Data

```{r enhanced-crosstab, eval=FALSE}
# In jamovi:
# 1. Select Analyses → Exploration → ClinicoPath Comparisons → Cross Tables
# 2. Put the exposure variable in "Variables" and the outcome in "Group Variable"
# 3. Choose "pcat" = chisq or fisher, and a "Table style" (sty):
#    nejm (default), lancet, hmisc, arsenal, finalfit or gtsummary
# 4. Tick "showSMD" for standardized mean differences
#
# Note: there is no "Enhanced Cross Table" analysis in ClinicoPathDescriptives, and
# Cross Tables does not compute odds ratios or risk estimates. For odds ratios with
# confidence intervals, use the Odds Ratio analysis in the jsurvival module
# (https://www.serdarbalci.com/jsurvival/).
```

## Example 5: Brain Cancer Geographic Distribution

```{r brain-cancer}
data("BrainCancerGeo_df")

# Geographic distribution analysis
str(BrainCancerGeo_df)
```

### Summary by Region

```{r geo-summary, eval=FALSE}
# In jamovi:
# 1. Select Analyses → Exploration → ClinicoPath Descriptives →
#    Summary of Continuous Variables
# 2. Add the numeric columns (here 'lat' and 'long') to "Variables"
# 3. Tick "Distribution diagnostics" and "Outlier detection" as needed
# 4. Describe the 'county' column with
#    Exploration → ClinicoPath Descriptives → Summary of Categorical Variables
#
# Note: there is no "Summary Data 2" analysis in ClinicoPathDescriptives, and
# Summary of Continuous Variables has no grouping option. BrainCancerGeo_df holds
# only county / lat / long, so no incidence rates are available in this dataset.
```

## Best Practices for Oncology Data Analysis

1. **Check Data Quality First**: Always run data quality checks before analysis
2. **Consider Clinical Relevance**: Group continuous variables using clinically meaningful cutoffs
3. **Report Missing Data**: Document patterns of missing data
4. **Use Appropriate Statistics**: Consider non-parametric tests for skewed distributions
5. **Multiple Comparisons**: Apply corrections when making multiple comparisons

## Integration with Clinical Workflows

### Example Workflow

```{r workflow, eval=FALSE}
# 1. Data Import and Cleaning
#    - Import from CSV/Excel
#    - Check for errors and inconsistencies

# 2. Initial Exploration
#    - Use Single Variable Quality Check / Multi-Variable Visual Quality
#      for quality assessment
#    - Generate Summary of Continuous Variables and Summary of
#      Categorical Variables reports

# 3. Comparative Analysis
#    - Create Table One for publication
#    - Perform cross-tabulations
#    - Calculate effect sizes

# 4. Export Results
#    - Copy tables to manuscript
#    - Export plots for presentations
```

## Advanced Features

### Chi-Square Post-hoc Tests with Weighted Data

For aggregated data (like the smoking-lung cancer contingency table):

```{r chisq-posthoc, eval=FALSE}
# In jamovi:
# 1. Select Analyses → Exploration → ClinicoPath Comparisons →
#    Chi-Square Post-Hoc Tests
# 2. Set "rows" and "cols"; for pre-aggregated tables put the frequency column
#    in "counts"
# 3. Tick "showResiduals" to identify the cells driving the association
#    (with "residualsCutoff", default 2)
# 4. "posthoc" is bonferroni by default; holm, fdr and none are also available.
#    "sig" (default 0.05) sets the significance threshold, and "testSelection"
#    (auto by default) chooses between chisquare and fisher
```

## Conclusion

The ClinicoPath Descriptives module provides comprehensive tools for analyzing oncology data. By combining these tools with the rich datasets from `OncoDataSets`, researchers can:

- Generate publication-ready descriptive statistics
- Identify patterns in cancer data
- Perform rigorous comparative analyses
- Ensure data quality throughout the analysis process

## References

- OncoDataSets package: https://cran.r-project.org/package=OncoDataSets
- ClinicoPathDescriptives documentation: https://www.serdarbalci.com/ClinicoPathDescriptives/